### 2020-05-22 serial.tex
a few changes to the notation

We illustrate changes in the forward and backward serial intervals over the course of an epidemic by applying it to a specific example of an epidemic model. 
We model disease spread with a renewal-equation model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, roberts2007model, champredon2018equivalence}.
Ignoring births and deaths, changes in the proportion of susceptible individuals $S(t)$ and incidence of infection $i(t)$ can be written as:
\begin{equation}
\begin{aligned}
\frac{\mathrm{d}S}{\mathrm{d}t} &= - i(t)\\
i(t) &= \Ro S(t) \int_0^\infty i(t-\tau) \gdist(\tau) \dtau,
\end{aligned}
\label{eq:renewal}
\end{equation}
where \Ro is the basic reproduction number, and $\gdist(\tau)$ is the intrinsic generation-interval distribution (i.e., the forward generation-interval distribution of a primary case in a population where the number of susceptibles remains constant).
Then, the forward generation-interval for a cohort of individuals that were infected at time $\psymp$ follows \citep{champredon2015intrinsic}:
\begin{equation}
\gdist_\psymp (\tau) \propto \gdist(\tau) S(\psymp + \tau),
\end{equation}
which allows us to separate the joint probability distribution $h_\psymp$ of the forward incubation period and the forward generation-interval distribution as a product of the proportion of susceptible individuals $S$ and the joint probability distribution $h$ of the forward incubation period and the intrinsic generation intervals:
\begin{equation}
h_\psymp (x, \tau) \propto h(x, \tau) S(\psymp + \tau).
\end{equation}
We further assume that the forward incubation period distributions does not vary across cohorts over the course of an epidemic, as they represent the life history of a disease, and use $k$ without subscript instead. 
Then, we have:
\begin{equation}
\begin{aligned}
\idist(x) &= \int_0^\infty h(x, \tau) \dtau\\
\gdist(\tau) &= \int_0^\infty h(x, \tau) \dx
\end{aligned}
\end{equation}
Finally, the case reproduction for this model is defined as follows:
\begin{equation}
\Rc(t) = \Ro \int_0^\infty \gdist(\tau) S(t+\tau) \dtau.
\end{equation}
The forward and backward serial-interval distributions are then calculated by substituting these quantities into \eref{fserial} and \eref{bserial}.
We use this framework to illustrate how the realized epidemiological time distributions vary over the course of an epidemic and depend on the perspective.

@@

We simulate the renewal equation model using a discrete-time approximation:
\begin{equation}
\begin{aligned}
i(t) &= \Ro S(t-\Delta t) \sum_{m=1}^{\ell} i(t-m \Delta t) \hat{\gdist}(m \Delta t) \\
S(t) &= S(t-\Delta t) - i(t)
\end{aligned}
\end{equation}
where $\hat{\gdist}$ is a discrete-time intrinsic generation-interval distribution that satisfies the following:
\begin{equation}
\hat{\gdist}(m \Delta t) = \frac{\gdist(m \Delta t)}{\sum_{i=1}^\ell \gdist(m \Delta t)}, \quad m=1, \dots, \ell
\end{equation}
The continuous-time intrinsic generation-interval distribution is parameterized using a log-normal distribution (Table 1). We define the forward incubation period distribution in a similar manner:
\begin{equation}
\hat{k}(m \Delta t) = \frac{k(m \Delta t)}{\sum_{i=1}^\ell k(m \Delta t)}, \quad m=1, \dots, \ell,
\end{equation}
where its continuous-time analog is also based on a log-normal distribution.
For brevity, we assume that the forward incubation periods and intrinsic generation intervals are independent:
\begin{equation}
\hat{h}(m \Delta t, \Delta t n) = \hat{k}(m \Delta t)\hat{\gdist}(\Delta t n), \quad m,n=1, \dots, \ell.
\end{equation}
We use $\Delta t = 0.025\,\textrm{days}$ and $\ell=2001$ for discretization steps.

@@

The intrinsic generation-interval distribution $\gdist(\tau)$ provides a link between $r$ and \Ro via the Euler-Lotka equation:
\begin{equation}
\frac{1}{\Ro} = \int_0^\infty \exp(-r\tau) \gdist(\tau) \dtau.
\end{equation}
Here, we prove that the initial forward serial-interval distribution $f_0$ also provides the same link:
\begin{equation}
\frac{1}{\Ro} = \int_{-\infty}^\infty \exp(-r\tau) f_{0}(\tau) \dtau,
\end{equation}
where the initial forward serial-interval distribution is defined as:
\begin{equation}
f_{0}(\tau) \propto \int_{-\infty}^{0} \int_{\alpha_1}^{\tau} \exp(r \alpha_1) h(-\alpha_1, \alpha_2 - \alpha_1) k(\tau - \alpha_2) \, \mathrm{d}\alpha_2\,\mathrm{d}\alpha_1.
\end{equation}

First, we rewrite the initial forward serial-interval distribution in the following form:
\begin{equation}
\begin{aligned}
f_{0}(\tau) &\propto \int_0^\infty \int_{-\pinf}^{\tau} \exp(-r\pinf) h(\pinf, \sinf + \pinf) k(\tau - \sinf) \mathrm{d}\sinf\,\mathrm{d}\pinf\\
&= \int_{-\infty}^{\tau} \int_{\max{(0,-\sinf)}}^{\infty} \exp(-r\pinf) h(\pinf, \sinf + \pinf)k(\tau - \sinf)\mathrm{d}\pinf\, \mathrm{d}\sinf\\
\end{aligned}
\end{equation}
It follows that $z(\sinf)$ describes the time between symptom onset of infector and infection of infectee:
\begin{equation}
z(\sinf) \propto \int_{\max{(0,-\sinf)}}^{\infty} \exp(-r\pinf) h(\pinf, \sinf + \pinf) \mathrm{d}\pinf.
\end{equation}
The initial forward serial-interval distribution $f_0$ can be expressed as a convolution of $z$ and $k$:
\begin{equation}
\begin{aligned}
f_{0}(\tau) &\propto \int_{-\infty}^{\tau} z(\sinf) k(\tau - \sinf) \mathrm{d}\sinf.\\
\end{aligned}
\end{equation}
Then, we have
\begin{equation}
\int_{-\infty}^\infty \exp(-r\tau) f_{0}(\tau) \mathrm{d} \tau = \int_{-\infty}^\infty \exp(-r\tau) z(\tau) \mathrm{d} \tau \int_{0}^\infty \exp(-r\tau) k(\tau) \mathrm{d} \tau
\end{equation}

Note that 
\begin{equation}
\begin{aligned}
&\int_{-\infty}^\infty \int_{\max{(0,-\sinf)}}^{\infty} \exp(-r\pinf) h(\pinf, \sinf + \pinf) \mathrm{d}\pinf \mathrm{d}\sinf\\
&= \int_{0}^\infty \int_{-\pinf}^\infty \exp(- r \pinf) h(\pinf, \sinf+\pinf) \mathrm{d}\sinf\,\mathrm{d} \pinf\\
&= \int_{0}^\infty \exp(- r \pinf) k(\pinf) d\pinf
\end{aligned}
\end{equation}
because $k$ is a marginal probability distribution of $h$.
Since $\int_{0}^\infty \exp(- r \pinf) k(\pinf) d\pinf$ is a normalization factor for probability distribution $z$, we have:
\begin{equation}
\int_{-\infty}^\infty \exp(-r\tau) z(\tau) \mathrm{d} \tau = \frac{\int_{-\infty}^\infty \exp(-r\tau) \int_{\max{(0,-\sinf)}}^{\infty} \exp(-r\pinf) h(\pinf, \sinf + \pinf) \mathrm{d}\pinf \mathrm{d}\tau}{\int_{0}^\infty \exp(- r \pinf) k(\pinf) d\pinf}.
\end{equation}
Then, we have:
\begin{equation}
\int_{-\infty}^\infty \exp(-r\tau) f_{0}(\tau) \mathrm{d} \tau = \int_{-\infty}^\infty \exp(-r\tau) \int_{\max{(0,-\sinf)}}^{\infty} \exp(-r\pinf) h(\pinf, \sinf + \pinf) \mathrm{d}\pinf \mathrm{d}\tau.
\end{equation}

>>>>

We illustrate changes in the forward and backward serial intervals over the course of an epidemic by applying it to a specific example of an epidemic model. 
We model disease spread with a renewal-equation model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, roberts2007model, champredon2018equivalence}.
Ignoring births and deaths, changes in the proportion of susceptible individuals $S(t)$ and incidence of infection $i(t)$ can be written as:
\begin{equation}
\begin{aligned}
\frac{\mathrm{d}S}{\mathrm{d}t} &= - i(t)\\
i(t) &= \Ro S(t) \int_0^\infty i(t-\tau) \gdist(\tau) \dtau,
\end{aligned}
\label{eq:renewal}
\end{equation}
where \Ro is the basic reproduction number, and $\gdist(\tau)$ is the intrinsic generation-interval distribution (i.e., the forward generation-interval distribution of a primary case in a population where the number of susceptibles remains constant).
Then, the forward generation-interval for a cohort of individuals that were infected at time $\psymp$ follows \citep{champredon2015intrinsic}:
\begin{equation}
\gdist_\psymp (\tau) \propto \gdist(\tau) S(\psymp + \tau),
\end{equation}
which allows us to separate the joint probability distribution $h_\psymp$ of the forward incubation period and the forward generation-interval distribution as a product of the proportion of susceptible individuals $S$ and the joint probability distribution $h$ of the forward incubation period and the intrinsic generation intervals:
\begin{equation}
h_\psymp (x, \tau) \propto h(x, \tau) S(\psymp + \tau).
\end{equation}
We further assume that the forward incubation period distributions does not vary across cohorts over the course of an epidemic, as they represent the life history of a disease, and use $\idist$ without subscript instead. 
Then, we have:
\begin{equation}
\begin{aligned}
\idist(x) &= \int_0^\infty h(x, \tau) \dtau\\
\gdist(\tau) &= \int_0^\infty h(x, \tau) \dx
\end{aligned}
\end{equation}
Finally, the case reproduction for this model is defined as follows:
\begin{equation}
\Rc(t) = \Ro \int_0^\infty \gdist(\tau) S(t+\tau) \dtau.
\end{equation}
The forward and backward serial-interval distributions are then calculated by substituting these quantities into \eref{fserial} and \eref{bserial}.
We use this framework to illustrate how the realized epidemiological time distributions vary over the course of an epidemic and depend on the perspective.

@@

We simulate the renewal equation model using a discrete-time approximation:
\begin{equation}
\begin{aligned}
i(t) &= \Ro S(t-\Delta t) \sum_{m=1}^{\ell} i(t-m \Delta t) \hat{\gdist}(m \Delta t) \\
S(t) &= S(t-\Delta t) - i(t)
\end{aligned}
\end{equation}
where $\hat{\gdist}$ is a discrete-time intrinsic generation-interval distribution that satisfies the following:
\begin{equation}
\hat{\gdist}(m \Delta t) = \frac{\gdist(m \Delta t)}{\sum_{i=1}^\ell \gdist(m \Delta t)}, \quad m=1, \dots, \ell
\end{equation}
The continuous-time intrinsic generation-interval distribution is parameterized using a log-normal distribution (Table 1). We define the intrinsic incubation period distribution in a similar manner:
\begin{equation}
\hat{\idist}(m \Delta t) = \frac{\idist(m \Delta t)}{\sum_{i=1}^\ell \idist(m \Delta t)}, \quad m=1, \dots, \ell,
\end{equation}
where its continuous-time analog is also based on a log-normal distribution.
For simplicity, we assume that the forward incubation periods and intrinsic generation intervals are independent:
\begin{equation}
\hat{h}(m \Delta t, \Delta t n) = \hat{\idist}(m \Delta t)\hat{\gdist}(\Delta t n), \quad m,n=1, \dots, \ell.
\end{equation}
We use $\Delta t = 0.025\,\textrm{days}$ and $\ell=2001$ for discretization steps.

@@

The intrinsic generation-interval distribution $\gdist(\tau)$ provides a link between $r$ and \Ro via the Euler-Lotka equation:
\begin{equation}
\frac{1}{\Ro} = \int_0^\infty \exp(-r\tau) \gdist(\tau) \dtau.
\end{equation}
Here, we prove that the initial forward serial-interval distribution $f_0$ also provides the same link:
\begin{equation}
\frac{1}{\Ro} = \int_{-\infty}^\infty \exp(-r\tau) f_{0}(\tau) \dtau,
\end{equation}
where the initial forward serial-interval distribution is defined as:
\begin{equation}
f_{0}(\tau) \propto \int_{-\infty}^{0} \int_{\alpha_1}^{\tau} \exp(r \alpha_1) h(-\alpha_1, \alpha_2 - \alpha_1) \idist(\tau - \alpha_2) \, \mathrm{d}\alpha_2\,\mathrm{d}\alpha_1.
\end{equation}

First, we rewrite the initial forward serial-interval distribution in the following form:
\begin{equation}
\begin{aligned}
f_{0}(\tau) &\propto \int_0^\infty \int_{-\pinf}^{\tau} \exp(-r\pinf) h(\pinf, \sinf + \pinf) \idist(\tau - \sinf) \mathrm{d}\sinf\,\mathrm{d}\pinf\\
&= \int_{-\infty}^{\tau} \int_{\max{(0,-\sinf)}}^{\infty} \exp(-r\pinf) h(\pinf, \sinf + \pinf)\idist(\tau - \sinf)\mathrm{d}\pinf\, \mathrm{d}\sinf\\
\end{aligned}
\end{equation}
It follows that $z(\sinf)$ describes the time between symptom onset of infector and infection of infectee:
\begin{equation}
z(\sinf) \propto \int_{\max{(0,-\sinf)}}^{\infty} \exp(-r\pinf) h(\pinf, \sinf + \pinf) \mathrm{d}\pinf.
\end{equation}
The initial forward serial-interval distribution $f_0$ can be expressed as a convolution of $z$ and $\idist$:
\begin{equation}
\begin{aligned}
f_{0}(\tau) &\propto \int_{-\infty}^{\tau} z(\sinf) \idist(\tau - \sinf) \mathrm{d}\sinf.\\
\end{aligned}
\end{equation}
Then, we have
\begin{equation}
\int_{-\infty}^\infty \exp(-r\tau) f_{0}(\tau) \mathrm{d} \tau = \int_{-\infty}^\infty \exp(-r\tau) z(\tau) \mathrm{d} \tau \int_{0}^\infty \exp(-r\tau) \idist(\tau) \mathrm{d} \tau
\end{equation}

Note that 
\begin{equation}
\begin{aligned}
&\int_{-\infty}^\infty \int_{\max{(0,-\sinf)}}^{\infty} \exp(-r\pinf) h(\pinf, \sinf + \pinf) \mathrm{d}\pinf \mathrm{d}\sinf\\
&= \int_{0}^\infty \int_{-\pinf}^\infty \exp(- r \pinf) h(\pinf, \sinf+\pinf) \mathrm{d}\sinf\,\mathrm{d} \pinf\\
&= \int_{0}^\infty \exp(- r \pinf) \idist(\pinf) d\pinf
\end{aligned}
\end{equation}
because $\idist$ is a marginal probability distribution of $h$.
Since $\int_{0}^\infty \exp(- r \pinf) \idist(\pinf) d\pinf$ is a normalization factor for probability distribution $z$, we have:
\begin{equation}
\int_{-\infty}^\infty \exp(-r\tau) z(\tau) \mathrm{d} \tau = \frac{\int_{-\infty}^\infty \exp(-r\tau) \int_{\max{(0,-\sinf)}}^{\infty} \exp(-r\pinf) h(\pinf, \sinf + \pinf) \mathrm{d}\pinf \mathrm{d}\tau}{\int_{0}^\infty \exp(- r \pinf) \idist(\pinf) d\pinf}.
\end{equation}
Then, we have:
\begin{equation}
\int_{-\infty}^\infty \exp(-r\tau) f_{0}(\tau) \mathrm{d} \tau = \int_{-\infty}^\infty \exp(-r\tau) \int_{\max{(0,-\sinf)}}^{\infty} \exp(-r\pinf) h(\pinf, \sinf + \pinf) \mathrm{d}\pinf \mathrm{d}\tau.
\end{equation}
